---
title: "RNA-seq Functional Enrichment"
subtitle: "Colorectal cancer VS Healthy"
author: "CBL Students | Ana Carolina Fernandes & Nuno Aguileira & Vitoria Correia"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 7
    fig-height: 6
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

# Study overview

NOTE: Write here the summary description of the study. Focused on the biological variables.

# Pipeline overview

IMAGE! (Look for inspiration here: <https://instructr.github.io/rnaseq_counts2bio/intro.html>)


# Differential Expression Analysis overview

Describe here! (do not forget to describe the design/model ~ disease_status)

# Setup and Load data

```{r}
#| label: setup
#| include: false

library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(magrittr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

# Load data
de_results <- readRDS("../data/processed/de_results.rds")
pca_results <- readRDS("../data/processed/pca_results.RDS")

```

# Get genes of interest

NOTE: TO DO | Annotate the procedure

```{r}
# Create a list to save the enrichment analysis results
fun_enrich <- list()

#
## Prepare list of significant DE genes in descending Log2FoldChange
#
fun_enrich$de_genes_fc <-
  de_results$sig |>
  dplyr::select(gene_id, log2FoldChange) |>
  dplyr::filter(abs(log2FoldChange) > 4) |>
  dplyr::arrange(dplyr::desc(log2FoldChange))

# Named fold change vector for all DE genes
fc_vector_all <- tibble::deframe(fun_enrich$de_genes_fc)

# Up regulated genes
fun_enrich$de_genes_up <-
  de_results$sig |>
  dplyr::select(gene_id, log2FoldChange) |>
  dplyr::filter(log2FoldChange > 2) |>
  dplyr::arrange(dplyr::desc(log2FoldChange))

# Named fold change vector for upregulated genes
fc_vector_up <- tibble::deframe(fun_enrich$de_genes_up)

# Down regulated genes
fun_enrich$de_genes_down <-
  de_results$sig |>
  dplyr::select(gene_id, log2FoldChange) |>
  dplyr::filter(log2FoldChange < -2) |>
  dplyr::arrange(dplyr::desc(log2FoldChange))

# Named fold change vector for downregulated genes
fc_vector_down <- tibble::deframe(fun_enrich$de_genes_down)

#
## Extract top genes driving PC1 (positive and negative loadings separately)
#
pca_loadings <- pca_results$pca_vst$rotation |>
  tibble::as_tibble(rownames = "gene_id") |>
  dplyr::mutate(gene_id = stringr::str_remove(gene_id, "\\.\\d+$")) |>
  dplyr::select(gene_id, PC1)

# Named vector of PC1 loadings (used in cnetplot and heatplot)
fc_vector_pc1 <- tibble::deframe(
  pca_loadings |> dplyr::arrange(dplyr::desc(PC1)))

# Define threshold for top 5% of loadings by absolute value
threshold_pc1 <- quantile(abs(pca_loadings$PC1), 0.95)

# Top 5% genes with positive PC1 loadings
pca_genes_pc1_pos <- pca_loadings |>
  dplyr::filter(PC1 > threshold_pc1) |>
  dplyr::pull(gene_id)

# Top 5% genes with negative PC1 loadings
pca_genes_pc1_neg <- pca_loadings |>
  dplyr::filter(PC1 < -threshold_pc1) |>
  dplyr::pull(gene_id)

# Check how many genes were selected
cat("Positive PC1 loadings (top 5%):", length(pca_genes_pc1_pos), "genes\n")
cat("Negative PC1 loadings (top 5%):", length(pca_genes_pc1_neg), "genes\n")

```


# Functional Enrichment

NOTE: TO DO | Annotate the Analysis! (Explain what is Functional Enrichment, and what was done)!

## GO Biological Process Ontology

### All genes | Up- and Down-regulated genes

```{r}

#
## Functional enrichment analysis
#

# Run GO enrichment analysis using the enrichGO function
fun_enrich$ego_all <- clusterProfiler::enrichGO(
  gene          = fun_enrich$de_genes_fc$gene_id,   # Genes of interest
  universe      = rownames(de_results$shrunk),      # Background gene set
  OrgDb         = org.Hs.eg.db,                     # Annotation database
  keyType       = "ENSEMBL",                        # Key type for gene identifiers
  readable      = TRUE,                             # Convert gene IDs to gene names
  ont           = "BP",                             # Ontology: can be "BP", "MF", "CC", or "ALL"
  pvalueCutoff  = 0.05,                             # P-value cutoff for significance
  qvalueCutoff  = 0.10                              # Q-value cutoff for significance
)

# Visualize the enriched GO terms
fun_enrich$dotplot <-
  enrichplot::dotplot(
    fun_enrich$ego_all,
    showCategory = 20,
    title = "GO BP | Enrichment barplot")

fun_enrich$heatplot <-
  enrichplot::heatplot(
    fun_enrich$ego_all,
    showCategory = 10,
    foldChange = fc_vector_all) +
  ggplot2::ggtitle("GO BP | Enrichment heatplot")

fun_enrich$emapplot <-
  enrichplot::emapplot(
    enrichplot::pairwise_termsim(fun_enrich$ego_all),
    showCategory = 15,
    layout = "nicely")

fun_enrich$cnetplot <-
  enrichplot::cnetplot(
    fun_enrich$ego_all,
    showCategory = 5,
    layout = "nicely",
    foldChange = fc_vector_all)

# Print plots
fun_enrich$dotplot
fun_enrich$heatplot
fun_enrich$emapplot
fun_enrich$cnetplot

```

### Up-regulated genes

```{r}

#
## Up regulated genes
#

# Run GO enrichment analysis using the enrichGO function
fun_enrich$ego_up <- clusterProfiler::enrichGO(
  gene          = fun_enrich$de_genes_up$gene_id,   # Genes of interest
  universe      = rownames(de_results$shrunk),      # Background gene set
  OrgDb         = org.Hs.eg.db,                     # Annotation database
  keyType       = "ENSEMBL",                        # Key type for gene identifiers
  readable      = TRUE,                             # Convert gene IDs to gene names
  ont           = "BP",                             # Ontology: can be "BP", "MF", "CC", or "ALL"
  pvalueCutoff  = 0.05,                             # P-value cutoff for significance
  qvalueCutoff  = 0.10                              # Q-value cutoff for significance
)

# Visualize the enriched GO terms
fun_enrich$dotplot_up <-
  enrichplot::dotplot(
    fun_enrich$ego_up,
    showCategory = 20,
    title = "GO BP | Enrichment barplot")

fun_enrich$heatplot_up <-
  enrichplot::heatplot(
    fun_enrich$ego_up,
    showCategory = 10,
    foldChange = fc_vector_up) +
  ggplot2::ggtitle("GO BP | Enrichment heatplot")

fun_enrich$emapplot_up <-
  enrichplot::emapplot(
    enrichplot::pairwise_termsim(fun_enrich$ego_up),
    showCategory = 15,
    layout = "nicely")

fun_enrich$cnetplot_up <-
  enrichplot::cnetplot(
    fun_enrich$ego_up,
    showCategory = 7,
    layout = "nicely",
    foldChange = fc_vector_up)

# Print plots
fun_enrich$dotplot_up
fun_enrich$heatplot_up
fun_enrich$emapplot_up
fun_enrich$cnetplot_up

```

### Down-regulated genes

```{r}

#
## Down regulated genes
#

# Run GO enrichment analysis using the enrichGO function
fun_enrich$ego_down <- clusterProfiler::enrichGO(
  gene          = fun_enrich$de_genes_down$gene_id,   # Genes of interest
  universe      = rownames(de_results$shrunk),        # Background gene set
  OrgDb         = org.Hs.eg.db,                       # Annotation database
  keyType       = "ENSEMBL",                          # Key type for gene identifiers
  readable      = TRUE,                               # Convert gene IDs to gene names
  ont           = "BP",                               # Ontology: can be "BP", "MF", "CC", or "ALL"
  pvalueCutoff  = 0.05,                               # P-value cutoff for significance
  qvalueCutoff  = 0.10                                # Q-value cutoff for significance
)

# Visualize the enriched GO terms
fun_enrich$dotplot_down <-
  enrichplot::dotplot(
    fun_enrich$ego_down,
    showCategory = 20,
    title = "GO BP | Enrichment barplot")

fun_enrich$heatplot_down <-
  enrichplot::heatplot(
    fun_enrich$ego_down,
    showCategory = 10,
    foldChange = fc_vector_down) +
  ggplot2::ggtitle("GO BP | Enrichment heatplot")

fun_enrich$emapplot_down <-
  enrichplot::emapplot(
    enrichplot::pairwise_termsim(fun_enrich$ego_down),
    showCategory = 15,
    layout = "nicely")

fun_enrich$cnetplot_down <-
  enrichplot::cnetplot(
    fun_enrich$ego_down,
    showCategory = 7,
    layout = "nicely",
    foldChange = fc_vector_down)

# Print plots
fun_enrich$dotplot_down
fun_enrich$heatplot_down
fun_enrich$emapplot_down
fun_enrich$cnetplot_down

```


### Visualize PC1 loadings

NOTE: Explain what are loadings, and why they matter.

#### Positive loadings

Genes with positive loadings are the ones that contribute mostly to cancer.

```{r}

#
## PC1 positive loadings
#

fun_enrich$ego_pc1_pos <- clusterProfiler::enrichGO(
  gene          = pca_genes_pc1_pos,
  universe      = pca_loadings$gene_id,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  readable      = TRUE,
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.10
)

# Visualize the enriched GO terms
fun_enrich$dotplot_pc1_pos <-
  enrichplot::dotplot(
    fun_enrich$ego_pc1_pos,
    showCategory = 20) +
  ggplot2::ggtitle("GO BP | PC1 positive loadings")

fun_enrich$emapplot_pc1_pos <-
  enrichplot::emapplot(
    enrichplot::pairwise_termsim(fun_enrich$ego_pc1_pos),
    showCategory = 15,
    layout = "nicely")

fun_enrich$cnetplot_pc1_pos <-
  enrichplot::cnetplot(
    fun_enrich$ego_pc1_pos,
    showCategory = 5,
    layout       = "nicely",
    foldChange   = fc_vector_pc1)

# Print plots
fun_enrich$dotplot_pc1_pos
fun_enrich$emapplot_pc1_pos
fun_enrich$cnetplot_pc1_pos

```

#### Negative loadings

Genes with negative loadings are the ones that contribute mostly to the healthy state (non-cancer).

```{r}
#
## PC1 negative loadings
#

fun_enrich$ego_pc1_neg <- clusterProfiler::enrichGO(
  gene          = pca_genes_pc1_neg,
  universe      = pca_loadings$gene_id,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  readable      = TRUE,
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.10
)

# Visualize the enriched GO terms
fun_enrich$dotplot_pc1_neg <-
  enrichplot::dotplot(
    fun_enrich$ego_pc1_neg,
    showCategory = 20) +
  ggplot2::ggtitle("GO BP | PC1 negative loadings")

fun_enrich$emapplot_pc1_neg <-
  enrichplot::emapplot(
    enrichplot::pairwise_termsim(fun_enrich$ego_pc1_neg),
    showCategory = 15,
    layout = "nicely")

fun_enrich$cnetplot_pc1_neg <-
  enrichplot::cnetplot(
    fun_enrich$ego_pc1_neg,
    showCategory = 5,
    layout       = "nicely",
    foldChange   = fc_vector_pc1)

# Print plots
fun_enrich$dotplot_pc1_neg
fun_enrich$emapplot_pc1_neg
fun_enrich$cnetplot_pc1_neg

```

## GO Molecular Function Ontology

